---
title: "Assessing Qualitative Individual Differences with Bayesian Hierarchical Latent-Mixture Models"
format: html
editor: visual
---

## Introduction

In this user guide, we introduce the R function `baid()` that can be used to assess the structure of individual differences in simple experimental tasks with Bayesian hierarchical latent-mixture models. Details of this analysis approach -- including complete model specification, derivation of posterior distributions, and Bayes factor calculations -- can be found in Schnuerch and Rouder (2025).

The function requires data from repeated trials per participant in two conditions. Responses are assumed to be continuous (e.g., RTs, ratings) and normally distributed. Let $Y_{ijk}$ denote the $k$th replicate for person $i$ in condition $j$, with $i = 1, \ldots, I$, $j = 1,2$, and $k = 1, \ldots, K_{ij}$. The model at trial level is given by $$Y_{ijk} \sim \mbox{Normal}(\alpha_i + x_j \beta_i, \sigma^2),$$ where $\alpha_i$ denotes a person-specific mean, $x_j = -.50, .50$ codes the condition, $\beta_i$ is the effect parameter of interest, and $\sigma^2$ denotes the trial-by-trial variance.

The function `baid()` fits a Bayesian hierarchical latent-mixture model with Markov chain Monte Carlo (MCMC). This model assumes $\beta_i$ to be distributed across individuals according to a three-class mixture model consisting of positive effects, $\beta_i > 0$, negative effect, $\beta_i < 0$, and null effects, $\beta_i = 0$. To assess the structure of individual differences, the user can specify different class structures (e.g., "everybody has a positive effect" vs. "everybody has a positive or a null effect" vs. "everybody has a positive or negative effect", etc.), and statistical evidence for the specified structures is assessed by means of Bayes factors. Additionally, the model estimates individual class probabilities (i.e., the probability for each individual to be in each of the three possible classes) and effect parameters. For details, see Schnuerch and Rouder (2025).

## Getting Started

Source the function from this repository to use it in `R`:

```{r}
#| echo: true

source("https://raw.githubusercontent.com/mschnuerch/baid/refs/heads/main/baid.R")
```

Required packages are automatically loaded or, if necessary, installed and loaded. After sourcing successfully, the function `baid()` will be available from the global environment and can be used in the following way:

```{r}
#| echo: true
#| eval: false

out <- baid(
  data,       # data frame; must contain columns sub, cond, y
  priors,     # list of priors
  n_samples,  # No of posterior samples per chain
  burnin,     # No of burnin samples to be discarded from the chains
  chains,     # No of chains 
  check_convergence, # logical; calculate convergence statistics?
  bayes_factors,     # logical; calculate Bayes factors?
  h1 =,       # hypothesized class structures to be tested against general model
  progress    # logical; display progress bar?
)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| results: asis

tabl.in <- "
| Arguments       |
|:-:|:------------------------------------------------------------------------|
| data | data frame containing the vectors 'sub' (participant ids), 'cond' (condition), and 'y' (dependent variable). See subsection *Data*.|
| priors | a named list containing prior settings (see subsection *Priors*). |
| n_samples | number of posterior samples per chain |
| burnin  | number of samples to be discarded before analysis (per chain) |
| chains | number of chains |
| check_covergence | a logical indicating whether convergence statistics should be calculated (default is true) |
| bayes_factors | a logical indicating whether Bayes factors for model comparison should be calculated (default is true)|
| h1 | a single numeric vector or a list of numeric vectors representing class structures to be tested against the general model (see subsection *Hypotheses*)
| progress | a logical indicating whether to show a progress bar (default is true)
"
cat(tabl.in) # output the table in a format good for HTML/PDF/docx conversion
```

### Data

The data must be provided as a data frame containing three columns:

-   `sub`: Participant identifiers for each trial
-   `cond`: Condition identifier; must be numeric and coded as `-.50` and `.50`
-   `y`: Dependent variable; must be numeric

### Priors

Priors must be provided as a named list containing the following elements:

-   `q`: a numeric vector providing the three shape parameters of the Dirichlet prior on $\boldsymbol \rho$ ($q_1, q_2, q_3$)
-   `mu`: a numeric vector providing mean and variance of normal prior on $\mu$
-   `nu1`: a numeric vector providing mean and variance of normal prior on $\nu_1$
-   `nu2`: a numeric vector providing mean and variance of normal prior on $\nu_2$
-   `eta2`: scale parameter $r_\eta$ of Inv-$\chi^2$ prior on $\eta^2$
-   `delta2_1`: scale parameter $r_1$ of Inv-$\chi^2$ prior on $\delta_1^2$
-   `delta2_2`: scale parameter $r_2$ of Inv-$\chi^2$ prior on $\delta_2^2$
-   `sigma2`: scale parameter $r_\sigma$ of Inv-$\chi^2$ prior on $\sigma^2$

### Hypotheses

To assess the global structure of individual differences, the user can specify different latent-class structures that will be compared to the general, three-class structure by means of Bayes factors. These structures must be provided as numeric vectors with three elements indicating for each class whether it is contained in the mixture (denoted by `1`) or not (`0`). The argument is either a single numeric vector or a list of numeric vectors.

Examples:

-   `c(1, 0, 0)`: A one-class structure where all individuals are in the positive class ("Everybody does")
-   `c(1, 0, 1)`: A two-class structure where all individuals are either in the positive or in the null class ("Some do, some don't")
-   `c(0, 1, 0)`\_ A one-class structure where all individuals are in the null class ("Nobody does")
-   `list(c(1, 0, 0), c(1, 0, 1), c(0, 1, 0))`: All of the above structures

## Output

The function returns a list with three elements:

-   `mcmc_samples`: A matrix containing all posterior samples for all parameters
-   `checks`: A matrix containing R-hat, bulk effective sample size, and tail effective sample size statistics for all parameters
-   `bf`: A list containing Bayes-factor model comparison results for each specified hypothesis

## Concluding Remarks

The R functions provided in this repository may be used for non-commercial purposes free of charge. Although considerable effort was put into developing and testing these functions, there is no warranty whatsoever. Please refer to Schnuerch and Rouder (2025) for further information on model development, prior specification, and implementation. We are grateful for comments, questions, or suggestions. Please address communication concerning `baid()` to [Martin Schnuerch](mailto:martin.schnuerch@uni-mannheim.de).
